Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-10-02 vertical-thermal-review.Rmd
minor clean-up

Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and are evidenced to play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005]. 
In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017b]. 
Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both classes of VOC are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein]. 
Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a]. 
However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021]. 
Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [@simpragaVerticalCanopyGradient2013]. 
This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. T
emperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008]. 
Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of VOCs  in forest thermal sensitivity across the vertical gradient. 
Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.

@@

Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
While woody growth rates, measured as radial stem growth, may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress], patterns are consistent when comparing growth rates across individuals within a closed-canopy forest: growth rate consistently increases with tree size [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015]. 
In contrast, in more open forests and for open-growth individual trees, growth rate declines with DHB [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress], pointing to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.

Vertical gradients also affect the climate sensitivity of metabolism and growth.
Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021]. 
This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`). 
More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower the capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020]. 
This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress].
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

@@

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From @anderson-teixeiraJointEffectsClimateinpress in press. In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020.](tree_rings/Figure_4_tree_rings.png)

>>>>

Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005]. 
In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017b]. 
Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both classes of VOC are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein]. 
Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a]. 
However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021]. 
Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [@simpragaVerticalCanopyGradient2013]. 
This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008]. 
Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of VOCs  in forest thermal sensitivity across the vertical gradient. 
Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.

@@

Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress].
In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015]. 
This points to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.

Vertical gradients also affect the climate sensitivity of metabolism and growth.
Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021]. 
This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`). 
More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce adjustment of traits related to carbon metabolism during drought [@bartholomewSmallTropicalForest2020]. 
This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have most commonly found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimateinpress], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress].
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

@@

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From @anderson-teixeiraJointEffectsClimateinpress. In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020.](tree_rings/Figure_4_tree_rings.png)
